---
title: "Analyzing the NYC Traffic Death Data"
author: "John Little"
date: '`r Sys.Date()`'
output:
  html_notebook: default
---

## Load R packages and the data

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(ggvis)

traffic <- read_csv('traffic-deaths.csv', col_types = list(time = col_character()))
```

```{r}
traffic
```

## Process/clean the data/display bar graph

```{r}
traffic %>% 
  mutate(vtype = if_else(is.na(victim_type), "unknown", victim_type)) %>%         # replace NA with "unknown"
  mutate(fvtype = factor(vtype, levels = names(sort(table(vtype))))) %>%          # duplicate victim_type as a factor vector
  mutate(timestamp = mdy_hms(paste(date, " ", time), tz="America/New_York")) %>%  # process time
  filter(!is.na(timestamp)) -> fulldate                                           # remove NAs and assign to new tibble

fulldate %>% 
  ggvis(~month(timestamp)) %>% 
  layer_bars()
```

## Visualizing variable separately for a quick summary

### Accidents over time

```{r accidents_by_month}

# In this plot, we want to summarize the data by month. The bar chart
# will count the number of accidents in each month; we just have to
# tell it how to find the month in the data.  It's easy to pull parts 
# of the timestamp out using lubridate functions like month(). 
# Type ?lubridate for information.

fulldate %>% 
  mutate(month_var = factor(month(timestamp), labels = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sept","Oct","Nov","Dec"))) %>% 
  ggvis(~month_var) %>% 
  layer_bars()


```

```{r accidents_by_hour, message=FALSE, warning=FALSE}

# This is similar to the previous plot, but now we're pulling out the
# hour of the timestamp.  What do you think factor() is doing here?
# Tip: try the same plot without the factor() command.

# ggplot(df.fulldate, aes(x=factor(hour(timestamp)))) + geom_bar()

```

```{r accidents_by_day}

# Try on your own: plot the number of accidents on different days of the week.

```

***

**Save a new version!**

Head back to the [README file](https://github.com/dukevis/mean-streets) for instructions on how to push a new version to your GitHub repository straight from RStudio.

***

### Categories of victims

```{r age_distribution}

# In previous plots, we used "geom_bar".  This chart is using "geom_histogram".
# Why might that be?

# ggplot(df.fulldate, aes(x=age)) + geom_histogram(bins=25)

```

```{r victims_by_gender}

# Here is the basic framework for a pie chart, but it could use a lot of help!  
# For starters, try:
#    - fixing the categories to change blank cells to "Unknown"
#    - combine "M" and "m"
#    - specify the order of the wedges in pie chart

# ggplot(df.fulldate, aes(x=factor(1),fill=gender)) + geom_bar(position="stack") + coord_polar(theta="y") 

```

## Looking for relationships between variables

```{r victims_by_hour}

# This is the same plot by hour we had above, but now the bars are colored
# according to the victim type.  The final line of the plot overrides the 
# default colors using part of a color palette from the RColorBrewer package.

ggplot(df.fulldate, aes(x=factor(hour(timestamp)))) + geom_bar(aes(fill=victim_type)) + scale_fill_manual(values=rev(brewer.pal(5,"Set2")))
```

```{r hour_by_dayofweek}

# This plot is often called a "heatmap". It compares the day of the week to the 
# hour of the day to see if there are cycles of accidents.  In ggplot2, heatmaps
# are called bin2d, or a chart with two dimensional binning.  This plot also
# using "theme_bw()" to apply a black and white background theme, instead of the
# grey background with white gridlines.

ggplot(df.fulldate, aes(x=hour(timestamp), y=wday(timestamp, label=TRUE))) + 
  geom_bin2d(binwidth=1) + 
  scale_fill_gradient(low="mistyrose",high="indianred4") + 
  theme_bw()

```

```{r accidents_by_day_a}

# What about a line chart that shows the number of accidents each day for the 
# entire year?  Well, it's a bit messy, but here's one way you might make that.

ggplot(df.fulldate, aes(timestamp)) + geom_freqpoly(binwidth=86400)
```

```{r accidents_by_day_b}

# Here's another way to make pretty much the same chart.  What are the differences?

ggplot(df.fulldate, aes(yday(timestamp))) + geom_freqpoly(binwidth=1)

```

```{r charges_by_victim}

# Now, what if we wanted to try to predict whether charges would be filed?
# This code does a few things:
#    - limit the data just to rows where "charges" is not empty
#    - create one bar for adult victims, one for child victims
#    - split the bar into two categories -- one where charges is "None", and 
#      another that includes everything else
#    - have the bars measure percentage instead of count

ggplot(df.fulldate[df.fulldate$charges != "",], aes(x=child_adult,fill=charges!="None")) + geom_bar(position="fill")

```

## Try on your own

* The charts need a **lot** of help with axis labels, colors, etc.  
    New to ggplot2? Check out the cheatsheet under Help --> Cheatsheets
* What other variables need cleaning?
* What other variables could be summarized or compared?

## Resources

* [R for Data Science, Hadley Wickham](http://r4ds.had.co.nz/)
* [Happy Git and GitHub for the useR](http://happygitwithr.com/)
* [Duke University Libraries' LearnR group](https://github.com/dukevis/learnr/wiki)
* [Data is Plural Datasets - regularly updated](https://docs.google.com/spreadsheets/d/1wZhPLMCHKJvwOkP4juclhjFgqIY8fQFMemwKL2c64vk/edit#gid=0)
* [Pro Git Book](https://git-scm.com/book/en/v2)
* [Introduction to ggplot2](http://rpubs.com/amz25/TVG-ggplot2)
